BOLD Parameter Estimation Using Particle Filters

### 2010-09-02 doc/thesis/particlefilters.tex
just need to edit discussion/conclusion, change captions and copy some graphs from defense

where $N_p$ is the number of particles or points used to describe the prior 
using a Mixture PDF. 
Note that any exponents will be explicitly labeled as such, to avoid confusion with
the particle numbering scheme.

@@

\begin{equation}
\alpha(X) \approx P(x_0) = \sum_{i=0}^{N_p} w^i\delta(X - x^i_0 ) dx
\end{equation}
where $\delta(x-x_0)$ is 1 if and only if $x = x_0$ 
(the Kronecker delta function).

@@

\subsection{Calculating Weights}
To calculate the weight of a particular particle, it is necessary to 
calculate both $q(x^i_{0:k} | y_{0:k})$ and $P(x^i_{0:k} | y_{0:k})$.
Note that $q(x^i_{0:k} | y_{0:k})$ may be simplified by assuming that 
$y_k$ doesn't contain any information about $x_{k-1}$. Technically this 
could be false; since later measurements may shed light on currently hidden
changes in $x$, however in this work it is a good assumption because the
input does not immediately effect the BOLD signal. 
\begin{equation}
q(x^i_{0:k} | y_{0:k}) = q(x^i_{0:k} | y_{0:k-1})
\label{eq:QAssump}
\end{equation}
The choice of the importance density is another design decision; however
it is common to use the integrated state equations. 
Although other importance density functions exist; for the particle filter
used here, the standard importance density will be used: the model
prior.
\begin{equation}
q(x_k | x_{k-1}, y_{0:k}) =  P(x_k | x_{k-1})
\label{eq:ImportanceDensity}
\end{equation}
The benefit of this choice for importance density is that an approximation for
$P(x_k | x_{k-1})$ is freely available; its simply the set of particles propagated
forward in time using the state equations. Additionally it makes
updating weights simple, as seen in \autoref{eq:weightevolve}.

@@

Additionally, for the particle filter $y_k$ and $y_{0:k-1}$ are 
constant across all particles, thus $P(y_k| y_{0:k-1})$ can
be dropped when the equality is changed to a proportion. 
Using these properties, $P(x^i_{0:k} | y_{0:k})$ may be broken up as follows 
(primarily using Bayes' Theorem):
\begin{equation}
\begin{array}{lclr}
P(x_{0:k} | y_{0:k}) & = & \frac{P(y_{0:k}, x_{0:k})}{P(y_{0:k})} & \\
 & = & \frac{P(y_k, x_{0:k} | y_{0:k-1}) \cancel{P(y_{0:k-1})}}{P(y_k | y_{0:k-1}) \cancel{P(y_{0:k-1})}} & \\
 & = & \frac{P(y_k| x_{0:k}, y_{0:k-1}) P(x_{0:k} | y_{0:k-1})}{P(y_k | y_{0:k-1}) } & \\
 & = & \frac{P(y_k| x_{0:k}, y_{0:k-1}) P(x_k | x_{0:k-1}, y_{0:k-1}) P(x_{0:k-1} | y_{0:k-1})}{P(y_k | y_{0:k-1})} &  \\
& = & \frac{P(y_k| x_k) P(x_k | x_{k-1}) P(x_{0:k-1} | y_{0:k-1})}{P(y_k | y_{0:k-1})}  & [\text{\autoref{eq:MarkovProperty}}]\\
& \propto & P(y_k| x_k) P(x_k | x_{k-1}) P(x_{0:k-1} | y_{0:k-1}) & [P(y_k|y_{0:k-1}) \text{ is constant}]
 \end{array}
 \label{eq:UpdateBayes}
\end{equation}
Plugging \autoref{eq:ImportanceDensity} and the result of \autoref{eq:UpdateBayes}
into \autoref{eq:weightfunc} leads to:
\begin{eqnarray}
w^i_k & \propto & \frac{P(y_k| x^i_k) \cancel{P(x^i_k | x^i_{k-1})} P(x^i_{0:k-1} | y_{0:k-1})}
                         {\cancel{P(x^i_k | x^i_{k-1})}q(x^i_{0:k-1} | y_{0:k-1})} \nonumber \\
& \propto & w^i_{k-1}P(y_k| x_k) 
\label{eq:weightevolve}
\end{eqnarray}

@@

\section{Sequential Importance Resampling}
\label{sec:Particle Filter Resampling}

\begin{figure}
\includegraphics[width=16cm]{images/particle_filter}
\caption{Particle Filter progression, note that the initial support is flat; the particles
are equally spaced between -10 and 10}
\end{figure}
As a consequence 
of the wide prior distribution (required for a proper discretization of a continuous
distribution), a significant proportion of particles will quickly
develop insignificant weights. 
While this does help
describe the tails of the distribution, it means a great deal of computation will be wasted.
Instead, it would be preferable if most of the computation is spent on the most probable regions.
Ideally the computation time spent on tails would be proportional to the actual size of the
tails. In this case particle locations would match the true posterior and all weights would
be equal.  The case where a large number of the weights have become extremely small
is called particle degeneracy. In  Lui et al. 
an ideal calculation of the effective number of particles is found based on the 
particles' true weight \cite{Liu1998b}. However, given that only an approximation 
for the true weight 
exists, they also provide a simple heuristic calculation of $N_{eff}$.
\begin{equation}
N_{eff} \approx \frac{\sum_{i=0}^{N_p} w_i}{\sum_{i=0}^{N_p} w_i^2}
\label{eq:neff}
\end{equation}
Unless the prior is particularly accurate,
$N_{eff}$ drops very quickly.  To alleviate particle degeneracy
a common technique known as resampling is often applied. The idea of resampling is to 
draw from the approximate posterior, thus generating a replica of the posterior with 
a better support. Therefore, a new set of particles may be drawn from the empirical
distribution as follows:
\begin{equation}
\hat{x}_j \sim \left(\sum_{i=0}^{N_p} w^i_k\delta(x - x^i_k)\right)
\end{equation}

>>>>

where $N_p$ is the number of particles or points used to describe the prior 
using a Mixture \ac{PDF}. 
Note that any exponents will be explicitly labeled as such, to avoid confusion with
the particle numbering scheme.

@@

\begin{equation}
\alpha(X) \approx P(x_0) = \int_{-\infty}^{\infty}
    \sum_{i=0}^{N_p} w^i\delta(X - x^i_0 ) dx
\end{equation}
where $\delta(x-x_0)$ is 1 if and only if $x = x_0$ 
(the Kronecker delta function).

@@

\subsection{Calculating Weights}
To calculate the weight of a particular particle, it is necessary to 
calculate both $q(x^i_{0:k} | y_{0:k})$ and $P(x^i_{0:k} | y_{0:k})$.
Note that $q(x^i_{0:k} | y_{0:k})$ may be simplified by assuming that 
$y_k$ doesn't contain any information about $x_{k-1}$. Technically this 
could be false; since later measurements may shed light on currently hidden
changes in $x$, however in this work it is a good assumption because the
input does not immediately effect the BOLD signal. 
\begin{equation}
q(x^i_{0:k} | y_{0:k}) = q(x^i_{0:k} | y_{0:k-1})
\label{eq:QAssump}
\end{equation}
The choice of the importance density is another design decision; however
it is common to use the integrated state equations. 
Although other importance density functions exist; for the particle filter
used here, the standard importance density will be used: the model
prior.
\begin{equation}
q(x_k | x_{k-1}, y_{0:k}) =  P(x_k | x_{k-1})
\label{eq:ImportanceDensity}
\end{equation}
This choice for importance density is conventient since its simply 
the set of particles propagated
forward in time using the state equations. Additionally it makes
updating weights simple, as seen later in \autoref{eq:weightevolve}.

@@

Additionally, for the particle filter $y_k$ and $y_{0:k-1}$ are 
constant across all particles, thus $P(y_k| y_{0:k-1})$ can
be dropped when the equality is changed to a proportion. 
Using these properties, $P(x^i_{0:k} | y_{0:k})$ may be broken up as follows 
(primarily using Bayes' Theorem):
\begin{equation}
\begin{array}{lclr}
P(x_{0:k} | y_{0:k}) & = & \frac{P(y_{0:k}, x_{0:k})}{P(y_{0:k})} & \\
 & = & \frac{P(y_k, x_{0:k} | y_{0:k-1}) \cancel{P(y_{0:k-1})}}{P(y_k | y_{0:k-1}) \cancel{P(y_{0:k-1})}} & \\
 & = & \frac{P(y_k| x_{0:k}, y_{0:k-1}) P(x_{0:k} | y_{0:k-1})}{P(y_k | y_{0:k-1}) } & \\
 & = & \frac{P(y_k| x_{0:k}, y_{0:k-1}) P(x_k | x_{0:k-1}, y_{0:k-1}) P(x_{0:k-1} | y_{0:k-1})}{P(y_k | y_{0:k-1})} &  \\
& = & \frac{P(y_k| x_k) P(x_k | x_{k-1}) P(x_{0:k-1} | y_{0:k-1})}{P(y_k | y_{0:k-1})}  & [\text{\autoref{eq:MarkovProperty}}]\\
& \propto & P(y_k| x_k) P(x_k | x_{k-1}) P(x_{0:k-1} | y_{0:k-1}) & [P(y_k|y_{0:k-1}) \text{ is constant}]
 \end{array}
 \label{eq:UpdateBayes}
\end{equation}
Inserting \autoref{eq:ImportanceDensity} and the result of \autoref{eq:UpdateBayes}
into \autoref{eq:weightfunc} leads to:
\begin{eqnarray}
w^i_k & \propto & \frac{P(y_k| x^i_k) \cancel{P(x^i_k | x^i_{k-1})} P(x^i_{0:k-1} | y_{0:k-1})}
                         {\cancel{P(x^i_k | x^i_{k-1})}q(x^i_{0:k-1} | y_{0:k-1})} \nonumber \\
& \propto & w^i_{k-1}P(y_k| x_k) 
\label{eq:weightevolve}
\end{eqnarray}

@@

\section{Sequential Importance Resampling}
\label{sec:Particle Filter Resampling}
\begin{figure}
\includegraphics[width=16cm]{images/particle_filter2}
\caption[Example Particle Filter Progression]{a) Initial particle locations
uniformly distributed in [-10, 10], with
weights set to give the State Prediction (green). When a measurement is
received it converted to a distribution approximating uncertainty. This
distribution is then used to alter the weights of the existing particles.
b) Another measurement is applied similar to a), c) Resampling draws
new particles in locations according to the Mixture \ac{PDF} approximated 
by the particle weights. d) Another measurement is applied as before,
but now with non-uniform particle locations.}
\end{figure}

As a consequence 
of the wide prior distribution (required for a proper discretization of a continuous
distribution), a significant proportion of particles will quickly
develop insignificant weights. 
While this does help
describe the tails of the distribution, it means a great deal of computation will be wasted.
Instead, it would be preferable if most of the computation is spent on the most probable regions.
Ideally the computation time spent on tails would be proportional to the actual size of the
tails. In this case particle locations would match the true posterior and all weights would
be equal.  The case where a large number of the weights have become extremely small
is called particle degeneracy. In  Lui et al. 
an ideal calculation of the effective number of particles is found based on the 
particles' true weight \cite{Liu1998b}. However, given that only an approximation 
for the true weight 
exists, they also provide a simple heuristic calculation of $N_{eff}$.
\begin{equation}
N_{eff} \approx \frac{\sum_{i=0}^{N_p} w_i}{\sum_{i=0}^{N_p} w_i^2}
\label{eq:neff}
\end{equation}
Unless the prior is particularly accurate,
$N_{eff}$ drops very quickly.  To alleviate particle degeneracy
a common technique known as resampling is often applied. The idea of resampling is to 
draw from the approximate posterior, thus generating a replica of the posterior with 
a better support. Therefore, a new set of particles may be drawn from the empirical
distribution as follows:
\begin{equation}
\hat{x}_j \sim \left(\sum_{i=0}^{N_p} w^i_k\delta(x - x^i_k)\right)
\end{equation}
